Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand

### 2015-04-24 draftfinalreport_v3.Rnw
update stuff

The numbers of cases per category and population sizes for the complete data set from 2007--2014 can be seen in \autoref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard errors in regression analyses. However, there are also numerous issues with the data for MLA ethnicity category above and beyond the small population sizes, creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}). The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}. Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).

@@

For all our statistical analyses (including those above not shown) we used a Poisson error structure with rate adjusted for population size, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{align*} \label{eq:reg}
 y &\sim \textsf{Poisson}(\lambda n)\\
 \log(\lambda) &= \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _s (x_s) + \beta _a{}_e(x_a \times x_e)
  \end{align*}
Where $y$ is the number of cases, $\lambda$ is cases per capita, $n$ is population size, $\alpha$ is the intercept, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity and $_s$ signifies NZDep.

@@

Plotting the lower level data from the area unit data, available from the NIR database, shows that there is heterogeneity at the local level and an obvious temporal lag, as the proportion vaccinated declines with birth cohort (\autoref{fig:mmr_nir}), though recent vaccination levels are consistently high for MMR1. For full details, see \autoref{fig:fig12006} (MMR1, 2006), \autoref{fig:fig22006} (MMR2, 2006), \autoref{fig:fig12007} (MMR1, 2007), \autoref{fig:fig22007} (MMR2, 2007), \autoref{fig:fig12008} (MMR1, 2008), \autoref{fig:fig22008} (MMR2, 2008), \autoref{fig:fig12009} (MMR1, 2009), \autoref{fig:fig22009} (MMR2, 2009), \autoref{fig:fig12010} (MMR1, 2010), \autoref{fig:fig22010} (MMR2, 2010), \autoref{fig:fig12011} (MMR1, 2011), \autoref{fig:fig22011} (MMR2, 2011), \autoref{fig:fig12012} (MMR1, 2012), \autoref{fig:fig22012} (MMR2, 2012), \autoref{fig:fig12013} (MMR1, 2013), \autoref{fig:fig22013} (MMR2, 2013), \autoref{fig:fig12014} (MMR1, 2014), \autoref{fig:fig22014} (MMR2, 2014), \autoref{table:tab_mmr_nir} and the supplementary files.

@@

The correlation between the percentage vaccinated against MMR1 and MMR2 across the area units is shown in figure \autoref{fig:cor_nir}.

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_cor.pdf}
     \end{center}
     \caption{The correlation between measles-mumps-rubella (MMR) vaccination doses (MMR1 and MMR2) at the area unit level from the National Immunisation Register database for 2006--2014.}
     \label{fig:cor_nir}
\end{figure}

Vaccination coverage (0--13 years) and serosurvey estimates of immunity ($>13$ years) among different age classes provided by the Ministry were also used to map the immunity among the area units for those ages not included on the NIR database. Vaccine efficacy for 0--4 year olds was not necessary for these analyses, for those $\leq$ 13 years 99\% vaccine efficacy was used (\autoref{table:sero}) and for $\geq$ 14 year olds equivocal serological results were considered non-immune (\autoref{table:sero}). Again there was heterogeneity at the local level (\autoref{fig:sero_imm_map}) with pockets of the nation and cities with lower coverage than other areas (\autoref{fig:sero_imm_a}, \autoref{fig:sero_imm_w}, and \autoref{fig:sero_imm_c}).

@@

\subsection{Measles importation risk methods}
\label{sub:imp_risk}
For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday. Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports. We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from. We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of measles importation. We use 0--19 year old data for the immigration figures, as no further details were available and it was determined this was more appropriate than using all age classes because > 19 year olds are believed to be immune to measles more or less globally. These age category data were not available for New Zealand travellers, so we use all data on all travellers. We present the data from 2013 because this year was the most recent with the relatively complete WHO measles data, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage. Data for 2013 are less complete than 2012, but substantially better than 2014.

@@

Globally, analyses of the most complete and recent year of data available, 2013, suggest measles incidence is highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}). Immigration by 0--19 year old non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia (\autoref{table:imtravel12} and \autoref{fig:immigration12}). A number of other countries then follow, including United Kingdom, China, Japan and the United States. Travel by New Zealanders is also dominated by that to Australia (\autoref{table:nztravel12} and \autoref{fig:nztravel12}). Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\autoref{table:immigration12} and \autoref{fig:immigration12}). However, many South Pacific nations also appear in the most visited country list by New Zealanders. Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2013 and immigration numbers of 0--19 year olds in 2013, suggest that those countries with greatest international travel, Australia, UK, China and South East Asian countries, pose the greatest risk to New Zealand for measles importation. Though immigration from and travel to some Asian countries is lower, some Asian countries also pose a high risk of measles importation to New Zealand. Note that South Pacific islands do not appear to pose a greater risk, despite large numbers of travellers of New Zealand origin, perhaps because of low measles incidence in these nations. These data are shown in \autoref{table:risk12}. The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in \autoref{table:nzrisk12}, \autoref{table:imrisk12}, \autoref{fig:nzrisk12} and \autoref{fig:imrisk12}. The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in \autoref{table:toprisk}.

@@

Though the global incidence of measles is declining, in recent years that decline has slowed (\autoref{fig:trendincidence}) and immigration rates to New Zealand have risen (\autoref{fig:travel}). This suggests that the risk of measles importation could stay at similar levels or increase, though further analyses are require to understand the interaction between these variables.  Of note, however, is the clear seasonality in immigration and travel (\autoref{fig:travel}).  This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation. Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling. The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (\autoref{fig:travel}).

@@

The additional data we received from the Ministry from the NIR database allows us to look at vaccination coverage as a finer scale. There is consistently high vaccination cover by MMR1 (often exceeding 95\%, see \autoref{fig:mmr_nir}), with lower rates for MMR2, approaching 90\% (\autoref{fig:mmr_nir}). The heterogeneity discussed above, however, does not appear to be the same for each area unit, as there is less correlation among the percent coverage for the two vaccines (\autoref{fig:cor_nir}) than might be expected, even at a yearly level (not shown).

The distribution of na\"{\i}ve among the DHBs also varies, though the proportion of na\"{\i}ve overall across DHBs is relatively similar (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).  We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB. However, it is worth bearing in mind that the data here assume that the serosurvey results are appropriate for all DHBs.

Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand. However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where additional effort and vigilance is required for measles vaccination and control efforts. The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}). Recently Australia was declared free of endemic measles and so the 2013 data must reflect imported measles there. Other interaction terms beyond the product of travel and infection incidence may be more appropriate but this will require substantial further analyses. Other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}). The WHO 2012 data were more complete than the WHO 2013 data, which suffers from data deficiencies, but due to the dynamic nature of measles epidemiology and consultation with the Ministry we present the WHO 2013 data. Lastly, more refined data on the age of travellers may improve the risk assessment, but these are currently unavailable.

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age.
\item European ethnicity comprise the largest number of measles cases.
\item Pacific people are statistically more at risk per capita of measles infection, primarily 0--2 year olds.
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
\item There is statistical support for 2--24 year old Pacific people being less at risk than European and Maori of the same age, as are 5--17 year old Asian children.
\item The majority of cases are unvaccinated.
\item The majority of vaccine failures occur in those people which received single vaccinations around 1 year old.
\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority in DHBs with larger urban areas, though the \%  na\"{\i}ve is similar across DHBs.
\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
\item There may be seasonal changes in risk of measles importation, with travel numbers peaking in December.
\end{itemize}

@@

To understand the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009. To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}. We are required to compute the generation time for measles to do so. The generation time is the average time an index case infects others after becoming infected. We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}. We then estimated $R_t$ from the notification data (confirmed and probable cases) for each outbreak, defining outbreaks in the dataset given their reported outbreak numbers. The geographic location of cases are shown in \autoref{fig:dhbcases}. The outbreaks we used in our analyses are shown in \autoref{fig:outbreaks}. Note these analyses do not use those sporadic cases not linked to other outbreaks.

@@

To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we use the well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population \citep{diekmann13}:
\begin{equation} \label{eq:eqn1}
\log\left(1-\Pe\right)+\Ro\Pe=0,
\end{equation}
where $\Ro$ is the basic reproduction number and $\Pe$ is the proportion of the population infected over the course of the outbreak.

If a proportion $x_0$ of the population is susceptible following vaccination, then the  reproduction number under vaccination is $\Rr_V=x_0\Ro$, and the final size equation becomes
\begin{equation} \label{eq:eqn2}
\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0.
\end{equation}

Hence the relationship between the proportion initially susceptible and the proportion infected in an epidemic is
\begin{equation} \label{eq:eqn3}
x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}.
\end{equation}

In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks, $\Pv$ is
\begin{equation} \label{eq:prop}
\Pv = x_0-1/\Ro.
\end{equation}

@@

The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}. Our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014, but for which data were available until 12 June 2014, a mean estimate of 3.66 (range 3.4--3.92). The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one. Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect outbreaks to occur if conditions remain the same. An important caveat to the analysis of the last outbreak is that because this outbreak that ended in August was ongoing when the data were collected (12 June 2014), and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious. However, this is not the only case for which the $R_v$ is greater than one (\autoref{fig:r0}).

These analyses also imply that the regular importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to increase immunisation to high enough levels that measles does not become endemic and large outbreaks prevented. One measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months (\autoref{fig:r0_2011}). This implies that outbreaks may persist within the population for a substantial period, even if $R_v$ values are not much greater than one. A caveat to this and other $R_v$ estimates is that the outbreaks may include some sporadic cases misclassified, but also exclude some sporadic cases as yet not associated with an outbreak. Undetected and unreported cases may lower the estimate. Thus the true basic reproduction numbers may differ to those estimated. The relative contributions of both to our estimates are currently unknown.

To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use \autoref{eq:prop}. The proportion of the na\"{i}ve population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (\autoref{table:attack}). These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (\autoref{sub:cost_benefit}). However, they require differing numbers of vaccinations per DHB. Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (\autoref{table:attack}). The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (\autoref{table:attack}). Note that there is also evidence for vaccine failure (\autoref{fig:vaccstat}) and pockets of susceptible populations \citep{glass4} may require greater vaccination effort than these estimates and this is considered further in the \autoref{sec:cb}.

The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (\autoref{fig:r0}). However, the na\"{\i}ve population (\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for some of the 2009--2014 outbreaks (\autoref{fig:r0}) suggests that catch up vaccination is necessary (\autoref{table:attack}). The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics. Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% currently na\"{\i}ve. Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, whereas \citep{glass4} showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult.

>>>>

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}. Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard error in regression analyses. However, there are also numerous issues with the data for MLA ethnicity category above and beyond the small population sizes, creating issues for estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}). The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}. Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).

@@

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

\begin{equation} \label{eq:reg}
 \log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a \times x_e)+ \log(population)  + \epsilon
  \end{equation}
Where $\alpha$ is the intercept, $y$ signifies cases, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity, $_n$ signifies NZDep, and $\epsilon$ is the error term.

@@

Plotting the lower level data from the 2013 area unit data, available from the NIR database, shows that there is heterogeneity at the local level and an obvious temporal lag, as the proportion vaccinated declines with birth cohort (\autoref{fig:mmr_nir}), though recent vaccination levels are consistently high for MMR1. For full details, see \autoref{fig:fig12006} (MMR1, 2006), \autoref{fig:fig22006} (MMR2, 2006), \autoref{fig:fig12007} (MMR1, 2007), \autoref{fig:fig22007} (MMR2, 2007), \autoref{fig:fig12008} (MMR1, 2008), \autoref{fig:fig22008} (MMR2, 2008), \autoref{fig:fig12009} (MMR1, 2009), \autoref{fig:fig22009} (MMR2, 2009), \autoref{fig:fig12010} (MMR1, 2010), \autoref{fig:fig22010} (MMR2, 2010), \autoref{fig:fig12011} (MMR1, 2011), \autoref{fig:fig22011} (MMR2, 2011), \autoref{fig:fig12012} (MMR1, 2012), \autoref{fig:fig22012} (MMR2, 2012), \autoref{fig:fig12013} (MMR1, 2013), \autoref{fig:fig22013} (MMR2, 2013), \autoref{fig:fig12014} (MMR1, 2014), \autoref{fig:fig22014} (MMR2, 2014), \autoref{table:tab_mmr_nir} and the supplementary files.

@@

The correlation between the percentage vaccinated against MMR1 and MMR2 for the 2004 2013 area units is shown in figure \autoref{fig:cor_nir}.

\begin{figure}[H]
     \begin{center}
     \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_cor.pdf}
     \end{center}
     \caption{The correlation between measles-mumps-rubella (MMR) vaccination doses (MMR1 and MMR2) at the 2013 area unit level from the National Immunisation Register database for 2006--2014.}
     \label{fig:cor_nir}
\end{figure}

Vaccination coverage (0--13 years) and serosurvey estimates of immunity ($>13$ years) among different age classes provided by the Ministry were also used to map the immunity among the 2004 2013 area units for those ages not included on the NIR database. Vaccine efficacy for 0--4 year olds was not necessary for these analyses, for those $\leq$ 13 years 99\% vaccine efficacy was used (\autoref{table:sero}) and for $\geq$ 14 year olds equivocal serological results were considered non-immune (\autoref{table:sero}). Again there was heterogeneity at the local level (\autoref{fig:sero_imm_map}) with pockets of the nation and cities with lower coverage than other areas (\autoref{fig:sero_imm_a}, \autoref{fig:sero_imm_w}, and \autoref{fig:sero_imm_c}).

@@

\subsection{Measles importation risk methods}
\label{sub:imp_risk}
For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday. Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports. We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from. We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of measles importation. We use 0--19 year old data for the immigration figures, as no further details were available and it was determined this was more appropriate than using all age classes because > 19 year olds are believe to be immune to measles more or less globally. These age category data were not available for New Zealand travellers, so we use all data on all travellers. We present the data from 2013 because this year was the most recent with the relatively complete WHO measles data, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage. Data for 2013 are less complete than 2012, but substantially better than 2014.

@@

Globally, analyses of the most complete and recent year of data available, 2013, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}). Immigration by 0--19 year old non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia \autoref{table:imtravel12} and \autoref{fig:immigration12}. A number of other countries then follow, including United Kingdom, China, Japan and the United States. Travel by New Zealanders is also dominated by that to Australia (\autoref{table:nztravel12} and \autoref{fig:nztravel12}). Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\autoref{table:immigration12} and \autoref{fig:immigration12}). However, many South Pacific nations also appear in the most visited country list by New Zealanders. Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2013 and immigration numbers of 0--19 year olds in 2013, suggest that those countries with greatest international travel, Australia, UK, China and South East Asian countries, pose the greatest risk to New Zealand for measles importation. Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand. Note the South Pacific island do not appear to pose a greater risk, despite large numbers of travellers of New Zealand origin, perhaps because of low measles incidence in these nations. These data are shown in \autoref{table:risk12}. The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in \autoref{table:nzrisk12} and \autoref{table:imrisk12} and \autoref{fig:nzrisk12} and \autoref{fig:imrisk12}. The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in \autoref{table:toprisk}.

@@

Though the global incidence of measles is declining, in recent years that decline has slowed (\autoref{fig:trendincidence}) and immigration rates to New Zealand have risen (\autoref{fig:travel}). This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.  Of note, however, is the clear seasonality in immigration and travel (\autoref{fig:travel}).  This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation. Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling. The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (\autoref{fig:travel}).

@@

The additional data we received from the Ministry from the NIR database allows us to look at vaccination coverage as a finer scale. There is consistently high vaccination cover by MMR1 (often exceeding 95\%, see \autoref{fig:mmr_nir}), with lower rates for MMR2, but apparently approaching 90\% (\autoref{fig:mmr_nir}). The heterogeneity discussed above, however, does not appear to be the same for each area unit, as there less correlation among the percent coverage for the two vaccines (\autoref{fig:cor_nir}) than might be expected, even at a yearly level (not shown).

The distribution of na\"{\i}ve among the DHBs also varies, though the proportion of na\"{\i}ve overall across DHBs is relatively similar (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).  We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB. However, it is worth bearing in mind that the data here assume that the serosurvey results are appropriate for all DHB.

Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand. However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilance is required for measles vaccination and control efforts. The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}). Recently Australia was declared free of endemic measles and so the 2013 data must reflect imported measles there. Other interaction terms beyond the product of travel and infection incidence may be more appropriate but this will require substantial further analyses. Other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}). The WHO 2012 data were more complete than the WHO 2013 data, which suffers from data deficiencies, but due to the dynamic nature of measles epidemiology and consultation with the Ministry we present the WHO 2013 data. Lastly, more refined data on the age of travellers may improve the risk assessment, but these are currently unavailable.

\subsection{Risk analysis summary}
\begin{itemize}
\item Risk of measles infection decreases significantly with age.
\item European ethnicity comprise the largest number of measles cases.
\item Pacific people are statistically more at risk per capita of measles infection.
\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
\item There is statistical support for 2--24 year old Pacific people being less at risk than European and Maori of the same age, as are 5--17 year old Asian children.
\item The majority of cases are unvaccinated.
\item The majority of vaccine failures occur in those people which received single vaccinations around 1 year old.
\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas, though the \%  na\"{\i}ve is similar across DHBs.
\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
\item There may be seasonal changes in risk of measles importation, with travel numbers peaking in December.
\end{itemize}

@@

To understand the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009. To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}. We are required to compute the generation time for measles to do so. The generation time is the average time an index case infects others after becoming infected. We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}. We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their reported outbreak numbers. The geographic locations of cases are shown in \autoref{fig:dhbcases}. The outbreaks we used in our analyses are shown in \autoref{fig:outbreaks}. Note these analyses do not use those sporadic cases not linked to other outbreaks.

@@

To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we use the well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population \citep{diekmann13}:

\begin{equation} \label{eq:eqn1}
\log\left(1-\Pe\right)+\Ro\Pe=0
\end{equation}

where $\Ro$ is the basic reproduction number and $\Pe$ is the proportion of the population infected over the course of the outbreak.

If a proportion $x_0$ of the population is susceptible following vaccination, then the  reproduction number under vaccination is $\Rr_V=x_0\Ro$, and the final size equation becomes

\begin{equation} \label{eq:eqn2}
\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0
\end{equation}

Hence the relationship between the proportion initially susceptible and the proportion infected in an epidemic is

\begin{equation} \label{eq:eqn3}
x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}
\end{equation}

In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks, $\Pv$ is

\begin{equation} \label{eq:prop}
\Pv = x_0-1/\Ro
\end{equation}

@@

The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}. Our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014, but for which data were available until 12 June 2014, a mean estimate of 3.66 (range 3.4--3.92). The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one. Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect outbreaks to occur if conditions remain the same. An important caveat to the analysis of the last outbreak is that because this outbreak that ended in August was ongoing when the data were collected, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious. However, this is not the only case for which the $R_v$ is greater than one (\autoref{fig:r0}).

These analyses also imply that the regular importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to increase immunisation to high enough levels that measles does not become endemic and large outbreaks prevented. One measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months (\autoref{fig:r0_2011}). This implies that outbreaks may persist within the population for a substantial period, even if $R_v$ values are not much greater than one. A caveat to this and other $R_v$ estimates is that the outbreaks may include some sporadic cases misclassified, but also exclude some sporadic cases as yet not associated with another outbreak. Undetected and unreported cases may lower the estimate. Thus the true basic reproduction numbers may differ to those estimated. The relative contributions of both to our estimates are currently unknown.

To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use \autoref{eq:prop}. The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (\autoref{table:attack}). These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (\autoref{sub:cost_benefit}). However, they require differing numbers of vaccinations per DHB. Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (\autoref{table:attack}). The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (\autoref{table:attack}). Note that there is also evidence for vaccine failure (\autoref{fig:vaccstat}) and pockets of susceptible populations \citep{glass4} may require greater vaccination effort than these estimates and this is considered further in the \autoref{sec:cb}.

The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (\autoref{fig:r0}). However, the na\"{\i}ve population (\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for some of the 2009--2014 outbreaks (\autoref{fig:r0}) suggests that catch up vaccination is necessary (\autoref{table:attack}). The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics. Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% currently na\"{\i}ve. Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, whereas \citep{glass4} showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult. Future analyses at a smaller, more local level would be useful, but the lack of appropriate data matching between the NIR data and the census unit area data currently prevent these.
